---
title: "Assignment 01 Solutions"
format: html
---

## For Your Revision
In order to facilitate you thinking about your assignment, I'm providing how I would approach each of the tasks listed in the assignment. This should allow you to go in 

```{r ldpkg}
library(tidyverse)

```
## Part I: Importing and inspecting

#### Tasks:

1. Use `list.files`, `read_`, and map to load the data into `R`.

```{r rdfils}

file_names <- list.files("/Users/mattwilliamson/Websites/isdr_assignments/assignment_01_org/data/original/", pattern = ".csv", full.names = TRUE)

purple_air_list <- file_names %>% 
  map(., read_csv)


```
[This one should be pretty straightforward. The only tricky bits are that I set the `pattern` argument to `".csv"` to make sure that I'm getting the right files for `read_csv` (which in this case is less important because that's all that's in the folder). I also set the `full.names` argument to `TRUE` to avoid issues when I render this file.]{style="color: blue;"}

2. Use `dim()`, `names()`, `glimpse()`, and `summary()` to explore the datasets.

```{r chkdta}
purple_air_list %>% 
  map(., dim)

purple_air_list %>% 
  walk(., glimpse)

purple_air_list %>% 
  map(., \(x) summary(x))
```
[This one should also be pretty straightforward. One thing you'll notice is that I used `walk` for `glimpse`, but had to use `map` for the other two. This is because `glimpse` is part of the `tidyverse` and so was designed to be called in a pipeline where the primary interest is not in saving the result to an object, but rather to get the _side effect_ (which is what `walk` is good for). The other two functions (`dim` and `summary`) are from base `R` and so they behave a little differently.]{style="color: blue;"}

3. Identify at least two potential data quality issues (e.g., missing values, inconsistent categories, unrealistic ranges). Explain why you think these data might be incorrect and how you might go about verifying them.

#### Reflection

1. What assumptions are you making about the data? Are they a reasonable sample? What biases that might exist? Do the potential errors you've spotted impact your analysis? 

## Part 2: Data Manipulation with tidyverse

#### Tasks

Use `dplyr` to:

1. Combine all of the datasets into a single dataframe 

2. Select variables relevant to the overarching research question.

3. Filter the data to observations from the last 2 years and create a new column that calculates the monthly change in pm2.5 (hint: you'll need `mutate`, `arrange`, and `lag`)

4. Group the data by year and state to calculate the annual mean and standard deviation of all numeric variables for each state.

5. Convert the ungrouped dataset from wide to long format. 

#### Reflection

1. How does tidyverse syntax support transparent and reproducible analysis?

2. In what ways could poor coding style or inconsistent workflows make this analysis harder for collaborators?

## Part 3: Functions and Iteration

#### Tasks

1. Write a function that filters your long-format data by a variable and generates a plot of that variable through time for each state. Do this in two steps. Write the function in pseudocode in this block:
```{r}
#| echo: true
#| eval: false


```

Then create the function in this block:
```{r}
#| echo: true

```

2. Create a vector of variables and use your function to generate a list of plots (using `map`).

3. Inspect your plots and choose the variable that follows the trends in pm2.5 the closest. Make a plot of pm2.5 and your chosen variable (one on the x axis and one on the y axis).

#### Reflection

1. Why is functional programming (functions + iteration) valuable in research workflows?

2. Compare this approach to manual repetition: what are the long-term costs/benefits?

## Part 4: Synthesis and Reflection

Write a short (200–300 word) reflection connecting this assignment to **your own graduate research**. You should reflect on the following questions:

* How do these skills apply to the kinds of data you will encounter?

* What challenges do you anticipate when implementing reproducible workflows in your field?

* How might these practices make your research more transparent, credible, and impactful?