---
title: "Pipes, Functions, and Iteration"
subtitle: "HES 505 Fall 2025: Session 6"
author: "Matt Williamson"
format: 
  revealjs:
    theme: mytheme.scss
    highlight-style: printing
    slide-number: true
    show-slide-number: print
    embed-resources: true  
---

## What is literate programming?

## Putting the fun in `function`

* What sorts of things do you often copy and paste?

* Has anything gone wrong?

* Is it fun to read?

## Anatomy of a function

* Arguments: What does the function need to know to run?

* Parameters: The values you supply to the arguments

* Body: The actual steps in the functions recipe

* Return: What does the function output?

## Anatomy of a function

```{r}
#| echo: true
#| eval: false

find_largest <- function(df, val, size){
  read data
  group by grp
  calculate sum(val) in groups
  filter top size obs
  return filtered df
}

```


## Anatomy of a function

```{r}
#| echo: true
#| eval: false

find_largest <- function(df, grp, val, size){
  df %>% #read data
  group_by({{ grp }}) %>% #group by grp
  summarize(total = sum({{ val }}, na.rm=TRUE) %>%  #calculate sum in groups
  slice_max(., n = size) #filter top size obs
  #return filtered df
}

```

## Helpful Hints: Functions

* Use `print("some helpful message")` to check intermediate steps in your function

* The `{{ }}` allows you to pass variables (not values)

* use explicit `return()` to return more than the last step of the function

## Documenting Functions

* What does this do?

* What arguments does it accept and what type should those arguments be?

* What does it return?

* See `roxygen2`

## Readable scripts: Pipes

* _Pipes_ are helpful when the output of a function _depends_ on something before it

* Prevents intermediate processing steps from flooding your environment 

* More readable then nested function calls

* Thinking about your functions...


## Helpful Hints: Pipes

* `|>` and ` %>%` pass _outputs_ of the left-hand as _inputs_ to the right-hand

* Default is first argument of right-hand

* `.` is the placeholder for ` %>% ` and `_` is the placeholder for `|>`

* `_` is much simpler and has less functionality

## Readable Documents: Iteration

* Doing the same thing over and over

* `for` loops are the classic use

```{r}
#| echo: true
#| eval: false 

long_vector_of_inputs
long_vector_of_outputs
for (i in 1:length(long_vector_of_inputs)){
  long_vector_of_outputs[i] <- do_something(arg1 = long_vector_of_inputs[i])
}

```

## Readable Documents: iteration

* `map` and `map_`: Do this function for each element of a vector or list

* Default returns a `list`; suffixes can change this (e.g., `map_dfr`)

```{r}
#| echo: true
#| eval: false 

long_list_of_outputs <- map(long_vector_of_inputs,
                            do_something)

```

## Readable Documents: iteration

* Also works with 1-time functions

```{r}
#| echo: true
#| eval: false 

long_list_of_outputs <- map(long_vector_of_inputs,
                            function(x) do_something(x) %>% 
                              do_something_else())

long_list_of_outputs <- map(long_vector_of_inputs,
                            \(x) do_something(x) %>% 
                              do_something_else())

```

## A Note on serial vs. parallel processing

* purrr still does things seriously

* furrr allows things to run in parallel

* Works when one step isn't dependent on the other


# Bringing it all together: Project Management

## Directory structure

## The `data` folder

## The `scripts` folder

## The `docs` folder

## Avoiding pitfalls with `git`

- Always pull before you start working on anything new

- Avoid committing large files

- Using `.gitignore`
