---
title: "Point Patterns II"
subtitle: "HES 505 Fall 2025: Session 16"
author: "Matt Williamson"
format: 
  revealjs:
    theme: mytheme.scss
    highlight-style: printing
    slide-number: true
    show-slide-number: print
    embed-resources: true  
---

```{r}
library(spatstat)

```


## Point Processes and Why You Should Care

* They're ubiquitous

* Well-specified to many social and ecological theories

* Helps build intuition for interpolation

## The Poisson Distribution

::: columns
::: {.column width="40%"}
* Discrete probability distribution

* Probability of events in a fixed interval of time or space

* Mean (expectation) = variance = $\lambda$

:::
::: {.column width="60%"}
```{r}
#| fig-width: 8
#| fig-height: 8
#| echo: false

hist(rpois(1000, 8), main = "Lambda = 8")
```
:::
:::

## Re-visiting CSR

::: columns
::: {.column width="70%"}
```{r}
#| fig-width: 8
#| fig-height: 8
#| echo: false


set.seed(2025)

W <- owin(c(0,1), c(0,1))

hpp <- rpoispp(lambda = 100, win = W)
plot(hpp, main = "Homogeneous Poisson Process (CSR)", asp=1)

```

:::
::: {.column width="30%"}

* Constant intensity 

* Points are independent

* Density is a crude estimate of intensity

:::
:::

## Revisting CSR

::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false
plot(density(hpp), main="Estimated Intensity (HPP)", asp = 1)

```

:::
::: {.column width="50%"}

```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false
Q.dim <- 10
Q  <- quadratcount(hpp, nx=Q.dim, ny=Q.dim)
A  <- spatstat.geom::area(hpp$window) / Q.dim^2 # Area for each quadrat
plot(Q / A, main="Density", col="grey20", entries = Q/A, asp = 1)

```
:::
:::

## First-Order Effects 

::: columns
::: {.column width="70%"}
```{r}
#| fig-width: 8
#| fig-height: 8
#| echo: false


lambda_fun <- function(x, y){ 300 * (x^2 + y^2) }
ipp <- rpoispp(lambda = lambda_fun, win = W)
plot(ipp, main = "Inhomogeneous Poisson Process", asp = 1)

```

:::
::: {.column width="30%"}

* Intensity varies spatially

* Points are independent

:::
:::

## First-Order Effects

::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false
plot(density(ipp), main="Estimated Intensity (IPP)", asp=1)

```

:::
::: {.column width="50%"}

```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false
Q.dim <- 10
Q  <- quadratcount(ipp, nx=Q.dim, ny=Q.dim)
A  <- spatstat.geom::area(ipp$window) / Q.dim^2 # Area for each quadrat
plot(Q / A, main="Density", col="grey20", entries = Q/A, asp=1)

```
:::
:::

## Second-Order Effects

::: columns
::: {.column width="70%"}
```{r}
lambda_parent <- function(x, y){ 30 * (x + y + 0.2) }
clustered <- rThomas(kappa = lambda_parent, sigma = 0.015, mu = 10, win = W)
plot(clustered, main = "Inhomogeneous Cluster Process")
```

:::
::: {.column width="30%"}

* Intensity of "parents" is first-order

* Intensity of "children" is second-order

:::
:::

## Second-Order Effects

::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false

plot(density(clustered), main="Estimated Intensity (Clustered)")

```

:::
::: {.column width="50%"}

```{r}
#| fig-width: 6
#| fig-height: 6
#| echo: false
Q.dim <- 10
Q  <- quadratcount(clustered, nx=Q.dim, ny=Q.dim)
A  <- spatstat.geom::area(clustered$window) / Q.dim^2 # Area for each quadrat
plot(Q / A, main="Density", col="grey20", entries = Q/A, asp=1)

```
:::
:::


## Key Points

* Poisson distribution: mean = variance = $\lambda$ -> First Order

* **Homogeneous** refers to the rate, not it's outcome

* Statistical tests ask "Could this outcome be generated by a constant rate?"

# Second-Order Analysis

## Distance based metrics

* Provide an estimate of the _second order_ effects

* _Mean nearest-neighbor distance_:
$$\hat{d}_{min} = \frac{\sum_{i = 1}^{m} d_{min}(\mathbf{s_i})}{n}$$

## Mean Nearest Neighbor Distance

::: columns
::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5

ANN_hpp <- apply(nndist(hpp, k=1:50),2,FUN=mean)
plot(ANN_hpp ~ eval(1:50), type="b", las=1, main = "ANN for Homogeneous PP")
```
:::
:::{.column width="50%"}

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 5

ANN_clust <- apply(nndist(clustered, k=1:50),2,FUN=mean)
plot(ANN_clust ~ eval(1:50), type="b", las=1, main = "ANN for Clustered PP")
```

:::
:::

## Ripley's $K$ Function

::: {style="font-size: 0.7em"}
* Nearest neighbor methods throw away a lot of information

* The _K_ function is an alternative, based on a series of circles with increasing radius

$$
\begin{equation}
K(d) = \lambda^{-1}E(N_d)
\end{equation}
$$

* We can test for clustering by comparing to the expectation:

$$
\begin{equation}
K_{CSR}(d) = \pi d^2
\end{equation}
$$

* if $k(d) > K_{CSR}(d)$ then there is clustering at the scale defined by $d$

:::

## Ripley's $K$ Function

* When working with a sample the distribution of $K$ is unknown

* Estimate with

$$
\begin{equation}
\hat{K}(d) = \hat{\lambda}^{-1}\sum_{i=1}^n\sum_{j=1}^n\frac{I(d_{ij} <d)}{n(n-1)}
\end{equation}
$$

where:

$$
\begin{equation}
\hat{\lambda} = \frac{n}{|A|}
\end{equation}
$$

## $K$ for HPP 

::: columns
::: {.column width="30%"}
```{r}
#| fig-width: 5
#| fig-height: 5
#| echo: false

plot(hpp, main = "Homogeneous Poisson Process (CSR)", asp=1)

```

:::
::: {.column width="70%"}

```{r}
#| fig-width: 9
#| fig-height: 7
#| echo: false

plot(Kest(hpp), main="K: Homogeneous Poisson")

```
:::
:::

## $K$ for IPP 

::: columns
::: {.column width="30%"}
```{r}
#| fig-width: 5
#| fig-height: 5
#| echo: false

plot(ipp, main = "Inhomogeneous Poisson Process (CSR)", asp=1)

```

:::
::: {.column width="70%"}

```{r}
#| fig-width: 9
#| fig-height: 7
#| echo: false

plot(Kest(ipp), main="K: Inhomogeneous Poisson")

```
:::
:::

## $K$ for Clustered Processes

::: columns
::: {.column width="30%"}
```{r}
#| fig-width: 5
#| fig-height: 5
#| echo: false

plot(clustered, main = "Clustered Poisson Process (CSR)", asp=1)

```

:::
::: {.column width="70%"}

```{r}
#| fig-width: 9
#| fig-height: 7
#| echo: false

plot(Kest(clustered), main="K: Clustered Poisson")

```
:::
:::

## Adjustments for $K$

* $K$ is sensitive to edge effects (window size) (`correction` argument to `Kest`)

* $K$ assumes HPP (`Kinhom` function instead of `Kest`)


## $K$ vs $Kinhom$ for IPP 

::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 5
#| fig-height: 7
#| echo: false

plot(Kest(ipp), main="K: Inhomogeneous Poisson")
```

:::
::: {.column width="50%"}

```{r}
#| fig-width: 5
#| fig-height: 7
#| echo: false

plot(Kinhom(ipp), main="Kinhom: Inhomogeneous Poisson")

```
:::
:::

## $K$ vs $Kinhom$ for Clustered PP

::: columns
::: {.column width="50%"}
```{r}
#| fig-width: 5
#| fig-height: 7
#| echo: false

plot(Kest(clustered), main="K: Clustered Poisson")
```

:::
::: {.column width="50%"}

```{r}
#| fig-width: 5
#| fig-height: 7
#| echo: false

plot(Kinhom(clustered), main="Kinhom: Clustered Poisson")

```
:::
:::


## Ripley's $K$ Function

* accounting for variation in $d$

```{r}
#| echo: false

kf.env <- envelope(clustered, correction="border", envelope = FALSE, verbose = FALSE)
plot(kf.env)
```

## Take-Aways

* Density and Intensity for first-order effects

* Distance for second-order effects

* Linkages for kriging and interpolation