---
title: "Tools of the Trade"
date: "2025-08-27"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Today we'll be talking about reproducible workflows, why they're important (generally) and some considerations for working with geographic data. We'll also get introduced to version control and it's the practice of open, reproducible research. Finally, you'll meet `Quarto` our go-to tool to try and keep all of the pieces of the data analysis pipeline together.


## Readings

The following readings are intended to give you some sense of the discussion surrounding the role of spatial data in understanding the world. They are a mix of old favorites and relatively recent reviews. You don't need to read all of them or memorize them, but they are worth a skim. I bet you'll find something interesting.

### Setting the Stage
- {{< fa arrow-up-right-from-square >}} [Open science, reproducibility, and transparency in ecology](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/eap.1822) by Powers and Hampton - discusses the importance of open science for ecologists.

- {{< fa arrow-up-right-from-square >}} [Practical Reproducibility in Geography and Geosciences](https://eartharxiv.org/repository/view/3216/) by Nilst and Pebesma - describes the importance of reproducibility for geospatial analysis.

- <i class="fas fa-book"></i> [The Whole Game](https://r4ds.hadley.nz/whole-game.html) - from Wickham et al., *R for Data Science* [@wickham2016r]. Focus on the sections that begin with "Workflow" to get a sense for how we'll start putting the pieces together.

- <i class="fas fa-book"></i>[The FAIR Guiding Principles for scientific data management and stewardship](https://www.nature.com/articles/sdata201618) by [-@wilkinson2016fair] provides an overview of the FAIR framework for accessible, reproducible data.

### Technical Details

- {{< fa arrow-up-right-from-square >}} [Authoring in Quarto](https://quarto.org/docs/get-started/authoring/rstudio.html) - an intro to Quarto for developing different kinds of documents. Lots of other resources linked here!!

### Coding Help

- {{< fa book >}} [Chapter 1 - 6](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf) in Venables et al., *An Introduction to R* [@venables2009introduction] - for a quick refresher on data types in R (it's only 30 pages)

- {{< fa arrow-up-right-from-square >}} [Chapters 1-2](https://intro2r.com/) in Douglas et al., *An Introduction to R* - provides another intro to R that's been updated and is an open-source book.

- {{< fa arrow-up-right-from-square >}} [This RStudio Education](https://education.rstudio.com/learn/beginner/) page has a lot of additional tutorials to help you get started with `R`.

## Objectives

By the end of today you should be able to:

* **Describe** the benefits of reproducible data analysis workflows

* **Explain** the benefits of version control

* **Generate** a `Quarto` document and render it into `.html`

* **Execute** your first `commit` in GitHub classroom

## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/02-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/02-slides.html"></iframe>
```

</div>