---
title: "Data Models, Coordinates, and Geometries"
date: "2025-09-15"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Now that you've gotten an introduction (or refresher) to the logic of tidy data and the `tidyverse`, we can move into the actual spatial portion of the class! The first step in doing that is understanding how/why spatial data differs from more traditional tabular datasets.  

## Readings

### Setting the Stage

[Data Types and Spatial Data Models](https://www.opengeomatics.ca/types-of-data.html) by Skeeter and Pickell provides a general overview of different data types and how we might think about storing them spatially.

{{< fa arrow-up-right-from-square >}} [Types of Spatial Data](https://www.paulamoraga.com/book-spatial/types-of-spatial-data.html) from Paula Moraga's new book Spatial Statistics for Data Science: Theory and Practice with `R` provides a nice overview of the types of spatial data from the perspective of a statistical analyst.

{{< fa arrow-up-right-from-square >}} [Attributes and Support](https://r-spatial.org/book/05-Attributes.html) from Pebesma and Bivand's Spatial Data Science with Applications in `R` gives more info and examples on the nature of the relationship between geometries and support.

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> The introductory vignette for the [`sf` package](https://r-spatial.github.io/sf/articles/sf1.html) has a lot of useful info on `sf` objects and conventions.

{{< fa arrow-up-right-from-square >}} [Chapter 2](https://r.geocompx.org/spatial-class) in Geocomputation with R [@lovelace_geocomputation_2019] provides and overview of using `sf` for vector datasets and `terra` for raster data.


## Objectives
By the end of today you should be able to:


* **Contrast** the different "views" of spatial data and their incorporation in GIS.

* **Identify** key elements that make data "spatial".

* **Recognize** the relationship between geometries and support.

* **Access** geographic attributes of spatial objects using `sf` and `terra`

## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/07-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/07-slides.html"></iframe>
```

</div>

