---
title: "Projections, Extent, and Resolution"
date: "2025-09-17"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Congratulations, You've loaded your first spatial data into `R`!! Now we're going to start learning how to access and manipulate the 3 fundamental properties of spatial data: coordinates/geometries, extent, and resolution.
We'll spend some time talking about why these things matter, both philosophically and operationally. Then learn how to use `sf` and `terra` to access and manipulate these attributes.

## Readings

### Setting the Stage

[Scale and Projections](https://open.lib.umn.edu/mapping/chapter/3-scale-and-projections/) by Matson et al. provides an accessible overview to what projections are and why we need them.

- <i class="fas fa-external-link-square-alt"></i> [Lying With Maps](https://open.lib.umn.edu/mapping/chapter/7-lying-with-maps/) by Nelson et al. provides a nice introduction to the ways that the representation of spatial data can be misleading. 

### Technical Details

{{< fa arrow-up-right-from-square >}} [Chapter 2](https://www.paulamoraga.com/book-spatial/spatial-data-in-r.html) from Moraga (2023) provides more explanation projections, coordinates, etc. and their representation in `R`.

- <i class="fas fa-external-link-square-alt"></i> The introductory vignette for the [`sf` package](https://r-spatial.github.io/sf/articles/sf1.html) has a lot of useful info on `sf` objects and conventions.

- <i class="fas fa-book"></i> [Section 2.2](https://geocompr.robinlovelace.net/spatial-class.html) on Vector Data and [Sections 5.1-5.3](https://geocompr.robinlovelace.net/geometric-operations.html) on Geographic Operations in Lovelace et al. [@lovelace_geocomputation_2019] - for more details about vectors and geometric operations on vectors.

- <i class="fas fa-external-link-square-alt"></i> [Chapter 2, Sections 1-3](https://r-spatial.org/book/02-Spaces.html) and [Chapter 3, Section 1](https://r-spatial.org/book/03-Geometries.html) of Spatial Data Science by Edzer Pebesma and Roger Bivand (of the `sf`, `sp`, `rgeos`, and `rgdal` packages)


## Objectives

By the end of today, you should be able to: 

* **Describe** a Coordinate Reference System and why they matter.

* **Understand** what a projection is, the constraints of projections, and how we navigate them in `R`

* **Link** concepts of scale to measurements like extent and resolution

* **Access** the projection, extent, and resolution using `R`

## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/08-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/08-slides.html"></iframe>
```

</div>
