---
title: "Attribute Data Operations"
date: "2025-09-22"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

At this point, you should be (somewhat) familiar with the `dplyr` approach to manipulating tabular data. No only that, but you can get spatial data into `R`!! Now it's time to start bringing different datasets together to answer our questions. For the next 3 sessions, we'll be working towards this. Today, the focus is on creating and manipulating attribute data, with a special focus on using spatial location as the foundation for doing that.

## Readings

### Setting the Stage

- <i class="fas fa-book"></i> [Spatio-temporal modeling of particulate air pollution in the conterminous United States using geographic and meteorological predictors](https://ehjournal.biomedcentral.com/articles/10.1186/1476-069X-13-63?utm_source=chatgpt.com#Sec2) by [@yanosky2014spatio] illustrates the use of combined vector, raster, and tabular data to generate high-resolution predictions of particulate matter exposure in the US. The authors use ArcGIS for the spatial stuff, but you could easily replicate this analysis in `R`.

- <i class="fas fa-book"></i> [Quantitative Conservation Geography](https://www.sciencedirect.com/science/article/pii/S0169534721002275?via%3Dihub#s0015) by [@didi2021conservation] highlights the role that geography plays in environmental issues, identifies the need to integrate datasets, but also makes some pretty big assumptions about the goals of conservation geographers and the applicability of the data at hand.

- <i class="fas fa-book"></i> [Manufacturing Desolation: Unauthorized Border Crosser Mortality, Disappearance, and the Sociopolitical Construction of Remoteness in U.S. Boundary Enforcement](https://www.tandfonline.com/doi/full/10.1080/24694452.2025.2514034?af=R#d1e282) by [@boyce2025manufacturing] is a really interesting example of combining social, environmental, and physiological data to understand how policy changes might alter human mortality through changes in the distribution of water.

### Technical Details

- <i class="fas fa-book"></i> [Attribute Data Operations](https://r.geocompx.org/attr) from [@lovelace_geocomputation_2019] links `tidyverse` and base `R` approaches to data manipulation to the `sf` world.

- <i class="fas fa-book"></i> [Welcome to `tidyterra`](https://dieghernan.github.io/tidyterra/articles/welcome.html) along with the rest of the articles on this page give a sense for the places where `tidyterra` can help integrate rasters into existing `dplyr` pipelines.



## Objectives

By the end of today, you should be able to: 

- **Define** Spatial Analysis and distinguish it from making maps. 

- **Recognize** the role of database structure and spatial geometry in generating attributes

- **Develop** appreciation for the link between attribute operations and analysis validity. 

- **Perform** attribute operations on vector and raster data.

## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/09-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/09-slides.html"></iframe>
```

</div>