---
title: "Geometry Data Operations and Troubleshooting"
date: "2025-09-29"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

This session and the next one will focus on giving you more practice with manipulating the geometries of `sf` objects, calculating new variables based on those geometries, and converting `sf` objects to raster objects for use in `terra`. The primary goal of these two sessions is to refine your ability to manipulate spatial data and combine them into the databases we'll need for conducting the analyeses in Part 3 of this course. 

## Readings
### Bigger Picture

* __None__ Use this as an opportunity to revisit (or read for the first time) some of the previous material that discusses inferential challenges with spatial data.

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> The introductory vignette for the [`sf` package](https://r-spatial.github.io/sf/articles/sf1.html) has a lot of useful info on `sf` objects and conventions.

- <i class="fas fa-book"></i> [Section 2.2](https://geocompr.robinlovelace.net/spatial-class.html) on Vector Data and [Sections 5.1-5.3](https://geocompr.robinlovelace.net/geometric-operations.html) on Geographic Operations in Lovelace et al. [@lovelace_geocomputation_2019] - for more details about vectors and geometric operations on vectors.

- <i class="fas fa-external-link-square-alt"></i> [Section 3.1 and 3.2](https://r-spatial.org/book/03-Geometries.html) of Spatial Data Science, a `bookdown` project by Edzer Pebesma and Roger Bivand (of the `sf`, `sp`, `rgeos`, and `rgdal` packages).

## Objectives

By the end of today, you should be able to: 

* **Use** transformers to fix or modify geometries

* **Calculate** new values using predicates.

* **Convert** vector data to raster data


## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/11-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/11-slides.html"></iframe>
```

</div>

