---
title: "Building Spatial Databases"
date: "2025-10-01"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

Now that you have the complete picture of predicates, measures, and transformers; it's time to use them on some actual data. This lecture is meant to be the "practical" application of the ideas you've learned in our previous discussions of manipulating vector and and raster data. We'll start by planning our analysis together and then working through the steps.

## Readings

### Technical Details

- <i class="fas fa-external-link-square-alt"></i> The introductory vignette for the [`sf` package](https://r-spatial.github.io/sf/articles/sf1.html) has a lot of useful info on `sf` objects and conventions.

- <i class="fas fa-book"></i> [Section 2.2](https://geocompr.robinlovelace.net/spatial-class.html) on Vector Data and [Sections 5.1-5.3](https://geocompr.robinlovelace.net/geometric-operations.html) on Geographic Operations in Lovelace et al. [@lovelace_geocomputation_2019] - for more details about vectors and geometric operations on vectors.

- <i class="fas fa-external-link-square-alt"></i> [Section 3.1 and 3.2](https://r-spatial.org/book/03-Geometries.html) of Spatial Data Science, a `bookdown` project by Edzer Pebesma and Roger Bivand (of the `sf`, `sp`, `rgeos`, and `rgdal` packages).

- <i class="fas fa-external-link-square-alt"></i> [Chapter 6 and 7](https://walker-data.com/census-r/mapping-census-data-with-r.html) of _Analyzing US Census Data: Methods, Maps and Models in R_ by Walker [-@walker2023] is a great reference for accessing Census data with the `tidycensus` and `tigris` packages. It's also a great introduction to building spatial databases for analysis.

## Objectives

By the end of today, you should be able to: 

* Practice pseudocode to plan an analysis

* Link geoprocessing steps to research questions

* Explore how geoprocessing choices affect results

## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/12-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/12-slides.html"></iframe>
```

</div>