---
title: "Point Patterns I"
date: "2025-10-13"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))

```

Point patterns, where the only data are the location of a set of points (and potentially a few attributes of the points), are among the simplest (in terms of data structure) and ubiquitous (in terms of their frequency in the real world) spatial data we will encounter. That said, statistical analysis of point patterns remains an area of rapid methodological development which continues to inform many areas of spatial statistical development.

## Resources

### Bigger Picture

- <i class="fas fa-external-link-square-alt"></i>  [Chapter 11: Point Pattern Analysis](https://mgimond.github.io/Spatial/chp11_0.html) in Manuel Gimond's Introduction to GIS and Spatial Analysis `bookdown` project provides a nice (and free) introduction to some of these introductory point process methods. 

- <i class="fas fa-external-link-square-alt"></i> [Rings, circles, and null-models for point pattern analysis in ecology](https://onlinelibrary.wiley.com/doi/full/10.1111/j.0030-1299.2004.12497.x?casa_token=D8RfesZ_AjUAAAAA%3AkLmb_B8BubHk4swR3flI9QS4sQnO87wZSImZXsQFD41Q9iEExMEO_bDTPvnvOHtERcit-3YmShhyU0g) by [@wiegand] provides an introduction to metrics for spatial clustering with applications in ecology.

- <i class="fas fa-external-link-square-alt"></i> [Ecological information from spatial patterns of plants: insights from point process theory](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2745.2009.01510.x) by [@law2009] provides a "plants-eye view" linkage between the sorts of the data ecologists often collect and the (somewhat esoteric) math of point patterns. Don't get intimidated by the math here, just try to get an intuition for how the density and distance metrics relate to theories of spatial processes.

- <i class="fas fa-external-link-square-alt"></i> [Spatial Point Pattern Analysis and Its Application in Geographical Epidemiology](https://www.jstor.org/stable/pdf/622936.pdf) by [@gatrell1996] provides a broad overview of point patterns and processes along with some extensions and critiques based on the nature of epidemiological data. Some important points here for anyone thinking about a process that might be considered "contagious" (or spreading). 

### Technical Details

- <i class="fas fa-book"></i> [Chapters 17 and 18](https://www.paulamoraga.com/book-spatial/spatial-point-patterns.html) on Spatial Point Processes and the `spatstat` package in Paula Moraga's book _Spatial Statistics for Data Science: Theory and Practice with R_.

- <i class="fas fa-book"></i> [Chapter 11](https://r-spatial.org/book/11-PointPattern.html) in [@sds] illustrates the interaction between the `sf` package and `spatstat` package for point pattern analysis.

## Objectives

By the end of today, you should be able to: 

- **Define** first, and second order effects in point patterns

- **Simulate** point patterns in `spatstat`

- **Compare** quadrat counts to kernel density estimates for describing point patterns



## Slides
The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/15-slides")
```

<div>

```{=html}
<iframe class="slide-deck" src="/slides/15-slides.html"></iframe>
```

</div>