---
title: "Iterating with Election Data"
---

## Objectives

Last time, we used some "brute force" approaches to cleaning up the election data. Lots of repetition and lots of copy-and-pasting. That's not ideal - it's not particularly readable (think about reading the same sentence over and over again with one word change) and it's pretty error-prone. We're going to clean that up by:

> Develop our own, custom functions and use iteration to eliminate redundancy.

You will still be working with the [FEC contribution receipts](https://www.fec.gov/data/) for our four member of Congress, but this time we'll do it in a more principled way.

## Getting Set Up

First, [join the repository](https://classroom.github.com/a/JFFqRu2D). Then, we'll have to [get our credentials](https://rstudio.aws.boisestate.edu/) and login to a [new RStudio session](https://rstudio.apps.aws.boisestate.edu/). Finally, create a new, version control, git-based project (using the link to the repo you just created). 

Once you've done that, you'll need to load your packages. In this case, it's just the `tidyverse`.

```{r ldpkg}
library(tidyverse)
```


## Reading in the Data

Last time, we copied the `read_delim` function 4x to read in all the data:

```{r rddata}
#| echo: true
#| eval: false
simpson_receipts <- read_delim("data/original/simpson_fec_17_26.csv",
                               delim = ",")

fulcher_receipts <- read_delim("data/original/fulcher_fec_17_26.csv",
                               delim = ",")

crapo_receipts <- read_delim("data/original/crapo_fec_17_26.csv",
                               delim = ",")

risch_receipts <- read_delim("data/original/risch_fec_17_26.csv",
                               delim = ",")

```

That's not ideal. We can do the same thing with `map` using two commands. First, we'll use `list.files()` to get a vector of filenames. Then, we'll pass that to `map`

```{r iterread}

contribution_files <- list.files(here::here("data/original/"), pattern = "17_26.csv", full.names = TRUE)

contribution_data <- map(contribution_files,
                         function(x) read_delim(x, delim=","))

```

I did a few things here. First, I used `here::here()` to make sure that `R` used the correct project directory when looking for my files. Then, I passed "17_26.csv" as the parameter for the `pattern` argument. The `pattern` argument excepts characters and regular expressions as a way to make sure that I only get the files that match our election data (this is why file naming is important!!). Finally, I passed `TRUE` to the `full.names` argument to ensure that I get the entire path (not just the filename). Once I've got the `contribution_files` I pass that to `map`. I had to write out the `function(x)` because I wanted to pass additional arguments to `read_delim`. 

You've probably noticed that there's an even more compact way to do this using a pipe:

```{r iterread2}

contribution_data <- list.files(here::here("data/original/"), 
                                 pattern = "17_26.csv", 
                                 full.names = TRUE) %>%  
  map(.,
                         function(x) read_delim(x, delim=","))

```

Better still, we can replace `read_delim` with `read_csv` and eliminate the need to pass additional arguments.

```{r iterread3}

contribution_data <- list.files(here::here("data/original/"), 
                                 pattern = "17_26.csv", 
                                 full.names = TRUE) %>%  
  map(., read_csv)

```
This is helpful, but I would like to keep track of who the data belongs to, so I'm going to step back and write my own 1-time function inside of `map`.

```{r iterread5}

contribution_data <- list.files(here::here("data/original/"), 
                                 pattern = "17_26.csv", 
                                 full.names = TRUE) %>%  
  map(~ read_csv(.x) %>%
    mutate(representative = str_extract(basename(.x), "^[^_]+(?=_fec_)"))
  )
```
I've added a `mutate` step to the pipeline to create a column that tracks the representative. The `str_extract` function is from `stringr` and relies on regular expressions to decide which characters to keep. This particular expression takes all of the characters before "_fec" in the filename. Regular expressions are often complicated and frustrating, but they can be extremely powerful. One of the places I rely on AI in my coding is to get these right!


Now, let's inspect the resulting object:

```{r iterinspect}
class(contribution_data)

```

This is the first time we've really encountered the `list` data class. Lists can hold multiple data types so long as the data exists in its own `slot`. So a list could have 1 slot that contains a vector, another slot that contains a matrix, and another that contains a dataframe. Lists are extremely flexible, but take a little getting used to.

One of the tricky bits is actually getting to data in each slot. We can do this by using the `[[` notation.

```{r iterinspect2}
glimpse(contribution_data[[1]])

```
By using `[[1]]` I'm telling `glimpse` to run on the object that's in the first slot of `contribution_data`. Once we're inside the `list` the data behaves the same way it would if it didn't exist in the list.

## Choosing the Correct Columns

Last time, we used a pretty clunky way to narrow down all of the data.

```{r lasttime}
#| echo: true
#| eval: false

select_vars <- c("is_individual", "contributor_name",  "contribution_receipt_amount", "committee_name", "report_year")

simpson_subset <- simpson_receipts %>% 
  select(all_of(select_vars)) %>% 
  filter(is_individual == TRUE)

fulcher_subset <- fulcher_receipts %>% 
  select(all_of(select_vars)) %>% 
  filter(is_individual == TRUE)

crapo_subset <- crapo_receipts %>% 
  select(all_of(select_vars)) %>% 
  filter(is_individual == TRUE)

risch_subset <- risch_receipts %>% 
  select(all_of(select_vars)) %>% 
  filter(is_individual == TRUE)

```

Let's clean this up creating our own function:

```{r newfun}

select_ind_donors <- function(df, vars, donor_type){
  df %>% 
    select(all_of({{ vars }})) %>% 
    filter(is_individual == donor_type)
}

```

Our function `select_ind_donors`, has a descriptive name and takes 3 arguments (`df`, `vars`, and `donor_type`). The one thing to notice is our use of _embracing_ to pass our variables into the function. Let's test it on 1 slot of our list.

```{r testfun}
select_vars <- c("is_individual", "contributor_name",  "contribution_receipt_amount", "committee_name", "report_year",
                 "representative")

test_fun <- select_ind_donors(df = contribution_data[[1]],
                              vars = select_vars,
                              donor_type = TRUE)
glimpse(test_fun)
```

Looks like that worked! Now let's use iteration to do that to all of the slots in our list.

```{r iterfun}
select_vars <- c("is_individual", "contributor_name",  "contribution_receipt_amount", "committee_name", "report_year",
                 "representative")

ind_donors <- map(contribution_data, 
                  \(x) select_ind_donors(df = x,
                              vars = select_vars,
                              donor_type = TRUE))

```

Here you can see the power of lists combined with `map`: each slot gets passed to our function so that we don't have to extract each dataframe to run our function.

The only problem left is to get this into a single dataframe. We can do that easily enough using `bind_rows`:

```{r iterfundf}
select_vars <- c("is_individual", "contributor_name",  "contribution_receipt_amount", "committee_name", "report_year",
                 "representative")

ind_donors <- map(contribution_data, 
                  \(x) select_ind_donors(df = x,
                              vars = select_vars,
                              donor_type = TRUE)) %>% 
  bind_rows()

```

## Final Thoughts

This is just the tip of the iceberg. We'll continue practicing building functions and iterating as we move throug the rest of the semester. For now, I just want you to get some practice building intution for how this works.