---
title: "Comparing Statistical Models"
---

```{r}
library(tidyverse)
library(terra)
library(geodata)
library(tidyterra)
library(caret)
```

## Getting Started 
This exercise is going to rely on simulated data so you don't need to join a repository. You'll just need to  [get our credentials](https://rstudio.aws.boisestate.edu/) and login to a [new RStudio session](https://rstudio.apps.aws.boisestate.edu/). I'm using simulation because one of the best ways to test out whether you've specified your model correctly is to simulate the process with _known coefficients_. If your statistical model returns those values, then, assuming you've simulated a reasonable generating process, you should be able to trust that the model you fit is actually doing what you want it to. 


## Get the Data

```{r}
elevation <- geodata::elevation_30s(country = "CRI", 
                                    path = tempdir())
climate <- geodata::worldclim_country(country = "CRI", 
                                      var = "bio", 
                                      res = 0.5, 
                                      path = tempdir())
names(climate) <- str_extract(names(climate), "bio_\\d+")
names(elevation) <- "elev"

climate_df <- as.data.frame(climate, na.rm=TRUE)

climate_subset <- climate[[sample(1:nlyr(climate),size = 8)]]
climate_aligned <- resample(climate_subset, elevation)

# Stack the predictors
predictors <- c(elevation, climate_aligned)


plot(predictors)

```
## Generate a sample

```{r}
set.seed(123)
n_points <- 750

# Sample random locations from the study area
sample_points <- spatSample(elevation, size = n_points, 
                           method = "random", na.rm = TRUE, 
                           xy = TRUE, values = FALSE)

# Extract environmental values at these points
env_values <- terra::extract(predictors, sample_points)

species_data <- tibble(
  x = sample_points[,1],
  y = sample_points[,2]) %>% 
  bind_cols(env_values) %>% 
  drop_na()  # Remove any NA values

species_data_scl <- species_data %>% 
  mutate(across(colnames(env_values), ~ as.numeric(scale(.x))))

```

## Specify predictors

```{r}
K <- nlyr(predictors)

gen_predictors <- function(K, style = c("random", "correlated"), K_corr = NULL, rho = NULL){
  if(style == "random"){
    preds <- runif(n=K, min = -3, max = 3)
  }else{
    if (is.null(K_corr) | is.null(rho)) {
      stop("You selected method = 'correlated', so you must supply values for K_corr and rho")
    }
    # build correlation matrix for correlated block
  Sigma <- matrix(rho, nrow = K_corr, ncol = K_corr)
  diag(Sigma) <- 1
  # generate correlated variables
  corr_preds <- MASS::mvrnorm(n = 1, mu = rep(0, K_corr), Sigma = Sigma)
  random_preds <- runif(n=(K-K_corr), min = -3, max = 3)
  preds <- sample(c(corr_preds, random_preds))
  }
return(preds)
} 

known_betas <- gen_predictors(K=9, style = "correlated", K_corr = 4, rho=0.7)

```

## Simulate Outcomes

```{r}

species_data_preds <- species_data_scl %>%
  mutate(
    across(
      all_of(colnames(env_values)),
      ~ .x * known_betas[which(colnames(env_values) == cur_column())]
    )
  ) %>% 
  mutate(. , linpred = rowSums(across(all_of(colnames(env_values)))),
         prob = plogis(linpred),
         presence = rbinom(n(), size = 1, prob = prob)) 


# Check the prevalence
species_data_preds %>%
  summarize(
    prevalence = mean(presence),
    n_present = sum(presence),
    n_absent = sum(presence == 0)
  )  
```

## Check Out Values

```{r}

ggplot() +
  geom_spatraster(data = elevation) +
  scale_fill_viridis_c(name = "Elevation (m)", alpha = 0.7) +
  geom_point(data = species_data_preds, 
             aes(x = x, y = y, color = factor(presence)),
             size = 1, alpha = 0.6) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"),
                     labels = c("Absent", "Present"),
                     name = "Species") +
  labs(title = "Species Distribution") +
  theme_minimal() +
  theme(legend.position = "right")

# Compare environmental values between presence and absence
# Reshape data for easier plotting
plot_data <- species_data_preds %>%
  select(all_of(c(colnames(env_values), "presence"))) %>%
  pivot_longer(cols = all_of(colnames(env_values)),
               names_to = "variable",
               values_to = "value") %>%
  mutate(
    presence = factor(presence, labels = c("Absent", "Present")),
    variable = factor(variable, 
                     levels = c(colnames(env_values)),
                     labels = c(colnames(env_values)))
  )

ggplot(plot_data, aes(x = presence, y = value, fill = presence)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Absent" = "blue", "Present" = "red")) +
  labs(title = "Environmental Conditions by Presence/Absence",
       x = "Species Presence",
       y = "Value") +
  theme_minimal() +
  theme(legend.position = "none")

```

## Fit Logistic Regression

```{r}

vars <- species_data_preds %>%
  select(colnames(env_values)) %>%
  names()

form <- reformulate(vars, response = "presence")
form



model <- glm(form ,
             data = species_data_preds,
             family = binomial(link = "logit"))

# Model summary
summary(model)

# Compare estimated coefficients to true values
tibble(
  Parameter = c("Intercept", colnames(env_values)),
  True_Value = c(0, known_betas),
  Estimated = coef(model),
  Difference = Estimated - True_Value,
  Pct_Error = (Difference / True_Value) * 100
)



```

## Fit Random Forest

```{r}

species_data_preds_rf <- species_data_preds %>% 
  mutate(presence = as.factor(presence))

rf_model <- caret::train(
    form,
    data = species_data_preds_rf,
    method = "ranger",
    importance = "impurity",
    tuneGrid = expand.grid(
        "mtry" = 4,
        "splitrule" = "gini",
        "min.node.size" = 5
    ),
    num.trees = 100
)

varimp_tbl <- rf_model$finalModel$variable.importance %>%
  enframe(name = "variable", value = "importance") %>%
  mutate(importance_scaled = importance / max(importance)) %>%
  arrange(desc(importance_scaled))

# View top variables
head(varimp_tbl, 10)

# Plot
ggplot(varimp_tbl, aes(x = reorder(variable, importance_scaled),
                       y = importance_scaled)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "Variable", y = "Scaled importance (0–1)",
       title = "Random Forest Variable Importance (Ranger)") +
  theme_minimal()

```

## Compare Model Implications

```{r}

scale_stack <- scale(predictors)
pred_map <- terra::predict(
  object = scale_stack,
  model = model,
  type = "response"
)


pred_map_rf <- terra::predict(
  object = scale_stack,
  model = rf_model$finalModel$forest,
  type = "response"
)

```